---
title: "Homework 3"
author: "Your Name Here"
date: 'Fall 2020'
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: paper
    highlight: tango
---

##### This homework is due by **1:30PM ET on Wednesday, November 18**.  

### Loading packages

The code chunk below appears in the Rmd file, but won't be displayed in your html output.  

```{r, include = FALSE}
library(tidyverse)
library(knitr)
```


### Problem 1: A few simple plots

For this problem we'll use the `diamonds` dataset from the `ggplot2` package.

##### (a) Base R graphics

Use the `qplot` function to create a histogram of `carat` with the bars colored `steelblue`.  
```{r}
qplot(diamonds$carat, fill = 'steelblue')
```

##### (b) qplot histogram

This time, use the `ggplot` function (not the `qplot` function!) to create a histogram of `depth`.  

```{r}
library(ggplot2)
ggplot(diamonds, aes(x=depth)) + geom_histogram()
```

##### (c) qplot violin plots

Use `qplot`  to create violin plots showing how `price` varies across diamond `cut`.  Specify `fill = cut` to get all the boxplots to be coloured differently.

```{r}
qplot(data = diamonds, cut, price, geom=c("violin"), fill = cut)
```

**Hint**: For this exercise, it will be useful to know that `violin` is a geometry (`geom`) built into `ggplot2`, and that `qplot` can be called by specifying the following arguments: qplot(x, y, data, geom, fill)

### Problem 2: ggplot practice 

For this exercise we'll go back to the Cars93 data set in the MASS library

##### (a) size mapping, geom_point()

Define a `ggplot` object using the Cars93 data set that you can use to view `Price` on the y-axis, `MPG.highway` on the x-axis, and set the `size` mapping to be based on `Horsepower`.  

Use `geom_point()` to create a scatterplot from your `ggplot` object.

```{r}
library(MASS)
plot = ggplot(data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower)) + geom_point()
plot
```

##### (b) colour mapping

Repeat part (a), this time also setting the `colour` mapping to be based on `Origin`.

```{r}
plot = ggplot(data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower, colour = Origin)) + geom_point()
plot
```

##### (c) changing color palette

Repeat part (b), this time using the `scale_colour_manual()` layer to specify that you want to use `cbPalette` as your color palette.

```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
plot = plot + scale_color_manual(values = cbPalette)
plot

# Edit me
```


### Problem 3: More ggplot2 practice

##### (a) stat_smooth()

Repeat part 2(b), this time using `stat_smooth()` to add a layer showing the smoothed curve representing how `Price` varies with `MPG.highway`.  

```{r}
plot = ggplot(data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower, colour = Origin)) + geom_point() + stat_smooth()
plot
```

##### (b) facet_grid() 

Use your ggplot object from 2(b) along with the `geom_point()` and `facet_grid` layers to create scatterplots of `Price` against `MPG.highway`, broken down by (conditioned on) `Origin`.  

```{r}
plot = ggplot(data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower, colour = Origin)) + geom_point() + facet_grid(Cars93$Origin)
plot
```

(Your code should produce a figure with two scatterplots, one for each level of `Origin`.)

##### (c) Overlaying regression lines.

Modify your solution to part (b) to also display regression lines for each scatterplot.

```{r}
plot = ggplot(data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower, colour = Origin)) + 
  geom_point() + facet_grid(Cars93$Origin) + geom_smooth(method = 'lm')
plot
```


### Problem 4: Visualizing data summaries

This problem uses the Adult dataset, which we load below.  The main variable of interest here is `high.income`, which indicates whether the individual's income was over $50K.  Anyone for whom `high.income == 1` is considered a "high earner".

```{r}
adult.data <- read_csv("http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data", 
                       trim_ws=TRUE, 
                       col_names=c("age", "type_employer", "fnlwgt", "education", "education_num","marital", 
                                   "occupation", "relationship", "race","sex", "capital_gain", "capital_loss", 
                                   "hr_per_week","country", "income"))

adult.data <- mutate(adult.data,
                     high.income = as.numeric(income == ">50K"))
```

##### (a) Income by education level

Use `dplyr` (tidyverse) functions to produce a summary table showing how many individuals there are in each `education` bin, and how the proportion of high earners varies across `education` levels.  Your output should be a tibble with column names: `education`, `count` and `high.earn.rate`.  

```{r}
educIncome <- adult.data %>% group_by(education) %>% summarize(count = n(), high.earn.rate = mean(high.income))
educIncome
```

##### (b) Constructing a bar chart

Using the `ggplot` and `geom_bar` commands along with your data summary from part **(a)** to create a bar chart showing the high earning rate on the y axis and `education` on the x axis.  Specify that the color of the bars should be determined by the number of individuals in each bin.

```{r}
educPlot <- ggplot(educIncome, aes(x = education, y = high.earn.rate, fill = count)) +
  geom_bar(stat='identity')
educPlot
```

##### (c) Improve upon your plot by following example 4 [from this guide](https://rpubs.com/sediaz/rotate_x_text) to rotate the x-axis text so that it doesn't overlap with itself.  (Any reasonable rotation is fine.  It doesn't have to be -90 degrees as illustrated in the linked example.)

```{r}
educPlot + theme(axis.text.x=element_text(angle =- 270, vjust = .6))
```

##### (d)  Further improve upon your plot from part **(c)** by using `mutate()` and `factor()` to change the ordering of the `education` variable labels to something that reflects *ascending order* of educational attainment.  You can use your discretion in how to break ties for the post-high school educational levels.

```{r}
educIncome <- educIncome %>% 
  mutate(education = factor(education, levels = c("Preschool", "1st-4th", "5th-6th", "7th-8th", "9th", "10th", "11th", "12th", "HS-grad","	
Prof-school", "Assoc-voc", "Assoc-acdm", "Some-college", "Bachelors", "Masters", "Doctorate")))

educPlot <- ggplot(educIncome, aes(x = education, y = high.earn.rate, fill = count)) +
  geom_bar(stat='identity') + theme(axis.text.x=element_text(angle =- 270, vjust = .6))

educPlot

```


##### (e) summary table with multiple splitting variables

Use `dplyr` functions to produce a summary table showing how the proportion of high earners varies across all combinations of the following variables: `sex`, `race`, and `marital` (marital status).  In addition to showing the proportion of high earners, your table should also show the number of individuals in each bin.  Your table should have column names: `sex`, `race`, `marital`, `count` and `high.earn.rate`.  

```{r}
marSexRace <- adult.data %>% group_by(sex, race, marital) %>% summarize(count = n(), high.earn.rate = mean(high.income))
marSexRace
```

##### (f) Nicer table output using `kable()`

Use the `kable()` function from the `knitr` library to display the table from part **(e)** in nice formatting.  You should use the `digits` argument to ensure that the values in your table are being rounded to a reasonable number of decimal places.  

```{r}
adult.data %>% group_by(sex, race, marital) %>% summarize(count = n(), high.earn.rate = mean(high.income)) %>%
  kable(format = 'markdown', digits = 3)
```


**Hint**: You will have to specify the appropriate `format` argument in the `kable()` function.  The notes for Lecture 6 have a working example of this.

### Problem 5: Getting the right plot

##### (a) A more complex bar chart.

Using the table you created in 4(e), use ggplot graphics to construct a plot that looks like [the one at this link](http://www.andrew.cmu.edu/user/achoulde/94842/homework/target_fig.png)


**Hint** You may find it useful to use the following layers: `facet_grid`, `coord_flip` (for horizontal bar charts), `theme` (rotating x axis text) and `guides` (removing fill legend). 

```{r, fig.height = 4, fig.width = 10}
ggplot(data = marSexRace, aes(x = high.earn.rate, y = marital, fill = race)) + geom_bar(stat='identity') +
  facet_grid(sex ~ race) + 
  guides(fill = 'none') + theme(axis.text.x=element_text(angle =- 270, vjust = .6)) +
  xlab("Proportion earning over $50K per year")+
  ylab("Marital Status")
  
```

##### (b)  Hiding code with `echo` 

Repeat part **(a)**, but this time set the `echo` argument of the code chunk in such a way that the code is not printed, but the plot is still displayed.

```{r, fig.height = 4, fig.width = 8, echo = FALSE}
ggplot(data = marSexRace, aes(x = high.earn.rate, y = marital, fill = race)) + geom_bar(stat='identity') +
  facet_grid(sex ~ race) + 
  guides(fill = 'none') + theme(axis.text.x=element_text(angle =- 270, vjust = .6)) +
  xlab("Proportion earning over $50K per year")+
  ylab("Marital Status")
```



