---
title: "94-842 Homework 1"
author: "Nolan Roosa"
date: 'Fall 2020'
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r}
library(tidyverse)
```

### Problem 1: Simple Boolean operations

> Tip:  Note that each of the code blocks in this Problem contain the expression `eval = FALSE`.  This tells R Markdown to display the code contained in the block, but not to evaluate it.  To check that your answer makes sense, be sure to try it out in the console with various choices of values for the variable `x`.   

> We have not covered functions yet, so you are **not** being asked to write functions for your solutions.   


##### (a) Checking equality.

Given a variable `x`, write a Boolean expression that will evaluate to `TRUE` if the variable `x` is equal to `95791` (the numeric value).  

```{r, eval = FALSE}
x == 95791
```

##### (b) Checking inequality.

Given a variable `x`, write a Boolean expression that will evaluate to `TRUE` if the variable `x` is *not* `NA` (i.e., is not missing).

```{r, eval = FALSE}
is.na(x)
```

##### (c) Checking if a number is in a given range. 

Given a (possibly negative) number `x`, write a Boolean expression that returns `TRUE` if and only if `x` is smaller than `-90` or bigger than `19`.  

```{r, eval=FALSE}
x < -90 | x > 19
```

##### (d) A more complicated example.

Given an integer number `x`, write a Boolean expression that returns `TRUE` if and only if `x` is an **even** number between 1 and 21 or 103 and 181.  

```{r, eval=FALSE}
x %% 2 == 0 & (between(x, 1, 21) | between(x, 103, 181))
```

**Tip**: Recall the modulus operator we saw in lecture 1: `%%`.  For integers `x` and `y`, `x %% y` is the remainder of `x` divided by `y`.  

### Problem 2: Vector Boolean operations

##### (a) R has two kinds of Boolean operators implemented, single (`&`, `|`) and double (`&&`, `||`).  

One of these operators takes advantage of something called *lazy evaluation* while the other does not.  They also don't behave the same way when applied to *vectors*.  

Read the help file (`help("||")`) and construct some examples to help figure out how the two behave.  

To help you get started, try out the following two examples in your console:

```{r, eval = FALSE}
# Example:  The variable y1 is never defined.  
# (Do not define it!)
# What happens when you run this code?
x1 <- 5
(x1 < 10) | (y1 > 2)
(x1 < 10) || (y1 > 2)
```

```{r, eval = FALSE}
# Define vectors
vec1 <- c(TRUE, FALSE, FALSE)
vec2 <- c(TRUE, TRUE, FALSE)

# Apply various Boolean operations to see what happens
vec1 & vec2
vec1 && vec2
vec1 | vec2
vec1 || vec2
```

Can you explain what's happening?  Write up a brief explanation below.

<font color="#336600">

|| is only evaluating the equality of the first element of the vector.
If the left most elements can be evaluated as true/false, that is what is outputted
and the following right elements are not evaluated. | is evaluating the whole vector 

</font>

##### (b) Using `all()`

Two people were asked to give their preferences between two options: [Facebook, Twitter], [Firefox, Chrome], [Mac, PC], [Summer, Winter].  Their results are given below.

```{r}
alice.prefs <- c("Twitter", "Chrome", "Mac", "Summer")
bob.prefs <- c("Facebook", "Chrome", "PC", "Summer")
```

Use the `all()` function to determine if the two people have identical preferences. (Your code should ouput a single Boolean value, either `TRUE` or `FALSE`)

```{r}
all(alice.prefs == bob.prefs)
```

##### (c) Using `any()` 

Use the `any()` function to determine if the two people have any preferences in common.  (Your code should output a single Boolean value, either `TRUE` or `FALSE`)
```{r}
any(alice.prefs == bob.prefs)
```



##### (d) Missing values.

Let `age` be the vector defined below.

```{r}
age <- c(18, NA, 25, 71, NA, 45, NA, NA, 18)
```

Write a Boolean expression that checks whether each entry of `age` is missing (recall missing values are denoted by `NA`).  Your expression should return a Boolean vector having the same length as `age`.

```{r}
is.na(age)
```

### Problem 3: Referencing vector elements

##### (a)  `which()` practice

Write code that returns the indexes of `age` that are missing (i.e., equal to NA).

```{r}
which(is.na(age))
```

##### (b) Getting non-missing values

Write code that uses [negative indexes](http://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture01/lecture01-94842.html#41) and your solution from (a) to return only the values of `age` that are *not* missing. (i.e., your code should result in a vector with elements: 18, 25, 71, 45, 18) 


```{r}
age[-which(is.na(age))]
```

##### (c)  A more direct way of getting non-missing values

Using the negation operator `!` and the `is.na()` function, write an expression that returns only the values of `age` that are *not* missing.

```{r}
age[which(!is.na(age))]
```

##### (d) More `which()` practice

For the next three problem we'll go back to the `cars` data set from Lab 1.

```{r}
speed <- cars$speed
dist <- cars$dist
```

Write code that returns the indexes of the cars that had a stopping distance of 12 feet or more.

```{r}
which(cars$dist > 12)
```

##### (e) New function: `which.max` 

Use the `which.max()` function to figure out which car had the *longest* stopping distance. (Your code should return the car's index.)

```{r}
which.max(cars$dist)
```

##### (f) New function: `which.min`

Use the `which.min()` function to figure out the *speed* of the car that had the *shortest* stopping distance.  (Your code should return the car's speed.)

```{r}
print(cars$speed[which.min(cars$dist)])
```

### Problem 4: Data frame basics


##### (a) Importing data.

In Lecture 2 we saw how to use the `read.table()` function to import the survey data.  Now we'll use a different function.  Use the `read.csv()` function to import the survey data into a variable called `survey`.

```{r}
survey <- read.csv("http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data2020.csv")
```

**Tip**: The data file is located at `http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data2020.csv`.  Do not download the file.  Import the data directly using the URL.

##### (b) `$` notation

Use the `$` operator to select the TVhours column from the `survey` data

```{r}
survey$TVhours
```

##### (c) [,] notation

Repeat part (b) using `[,]` notation.  i.e., Use `[,]` notation to select the TVhours column from the `survey` data by name (i.e., obtain this column by using the name "TVhours" instead of using the column number)

```{r}
survey[,"TVhours"]
```

##### (d) [[]] notation

Repeat part (c) with [[]] notation.

```{r}
survey[["TVhours"]]
```

##### (e) [] notation

Repeat part (d), but this time using single blackets (`[ ]`) notation.  

(Observe that this returns a new single-column *data frame*, not just a vector.) 

```{r}
survey["TVhours"]
```

##### (f) `filter()` and `select()` practice

Use the `filter()` and `select()` functions *with piping* to pull the Editor and TV watching data for all PPM students who have no prior programming experience (PriorExp).   

```{r}
survey %>%
  filter(Program == 'PPM', PriorExp == 'Never programmed before' ) %>%
  select(TVhours, Editor)
```


### Problem 5: Data summaries and inline code practice.

##### (a) Bar graph

Use `qplot` to create a bar graph of respondents' Rexperience.

```{r, fig.align='center', fig.width=7, fig.height=4}
qplot(survey$Rexperience)
```

##### (b) Inline code practice

Replace all occurrences of ???? in the paragraph below with an inline code chunk supplying the appropriate information.

> Of the `r nrow(survey)` survey respondents, `r sum(survey[["Program"]] != "PPM")` were NOT from the PPM program.  We found that `r sum(survey[["OperatingSystem"]] == "Windows") / sum(nrow(survey))`% of the all students in the class use the Windows operating system.  `r sum(survey[["Rexperience"]] == "Never used") / sum(survey[["Program"]] == "PPM")`% of of PPM students report having never used R.

